---
title: "Final Project"
output: pdf
---

```{r}
library(tidyverse)
library(dplyr)
```

I am working with data from https://www.kaggle.com/augustus0498/life-expectancy-who.

This dataset, from WHO, has data on the country level to better understand facotrs that influence life expectancy.

```{r}
path="/Users/Raphael/Desktop/Class/BIOS512/final_project"
setwd(path)

data<-read.csv("led.csv")
data %>% head
```

Clean the file to make sure no NAs exist.
```{r}
library(stats) # for complete.cases
sum(is.na(data))
data=data[complete.cases(data),]
dim(data)
```
Now we work off of the remaining 1649 samples.

We first begin by looking into the countries and their life expectancies.

```{r}
p = ggplot(data, aes(x=Country))
p = p + geom_bar(aes(y=Lifeexpectancy), stat='identity')
p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
p = p + labs(x = 'Country', y = 'Life Expectancy')
p


```

This is messy so let's filter to countries on the extremes, or those that are above and below a certain threshold.
```{r}
data <- 
```

Time based analysis
Facet by country.
